---
title: "Simulation-and-MLE-Statistical-Inference"
author: "KAZI ABDUL MUBIN"
date: "17-Oct-2025"
format: pdf
editor: visual
toc: true
number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

```

# Introduction

We study simulation-based inference for modelling completion times in a security-training setting. Three estimators — **mean**, **median**, and **10% trimmed mean** — are evaluated under (i) a clean exponential model and (ii) a contaminated model with extreme values due to logging errors. Using 1,000 Monte Carlo replications of size 250, we quantify **Bias**, **Monte Carlo SD (MCSD)**, and **Mean Squared Error (MSE)** and provide brief, evidence-based answers to the reflection questions.

# Section 1 — Estimating Typical Completion Time for Security Training

## Part A: Natural Variation in Completion Times

### Scenario

Completion times follow $X\sim \mathrm{Exp}(\lambda)$ with $\lambda=1$. We compare the **sample mean**, **median**, and **10% trimmed mean** as summaries of a typical completion time when the model holds.

### Task 1 — Generate 1,000 samples ($n=250$) from $\mathrm{Exp}(1)$

```{r}
set.seed(6191)
library(dplyr); library(tibble); library(knitr)

lambda    <- 1
true_mean <- 1/lambda
n         <- 250
n_sims    <- 1000

samples_A <- replicate(n_sims, rexp(n, rate = lambda), simplify = FALSE)
```

\newpage

### Task 2 — For each sample, compute mean, median, and 10% trimmed mean

```{r}
est_stats <- function(x) {
  c(
    mean   = mean(x),
    median = median(x),
    trimmed_mean = mean(x, trim = 0.10)
  )
}

estimates_A <- do.call(rbind, lapply(samples_A, est_stats)) |> as.data.frame()
head(estimates_A)

```

### Task 3 — Compute Bias, MCSD, and MSE

Definitions:

$$
\begin{aligned}
\text{Bias} &= \overline{\hat{\theta}} - \theta_0,\\
\text{MCSD} &= \operatorname{sd}(\hat{\theta}),\\
\text{MSE}  &= \mathbb{E}\!\left[(\hat{\theta}-\theta_0)^2\right].
\end{aligned}
$$

```{r}
summarise_mc <- function(df, theta0) {
  sapply(df, function(z) c(
    Bias = mean(z) - theta0,
    MCSD = sd(z),
    MSE  = mean((z - theta0)^2)
  )) |>
    t() |>
    as.data.frame() |>
    tibble::rownames_to_column("Estimator")
}

est_stats <- function(x) {
  c(
    mean   = mean(x),
    median = median(x),
    trimmed_mean = mean(x, trim = 0.10)
  )
}

mc_summary_A <- summarise_mc(estimates_A, true_mean)

knitr::kable(
  mc_summary_A,
  digits = 4,
  caption = "Part A — Bias, MCSD and MSE (Exp(1), n = 250, 1,000 simulations)"
)

```

### Task 4 — Summary Table

| Estimator | Bias (≈) | MCSD (≈) | MSE (≈) | Interpretation Summary |
|:--------------|:-------------:|:-------------:|:-------------:|:--------------|
| **Mean** | 0.000 | 0.062 | 0.0038 | Unbiased and most efficient under the model assumptions |
| **Median** | −0.305 | 0.061 | 0.0967 | Negatively biased because the exponential median (0.693) is below the mean |
| **Trimmed Mean (10%)** | −0.168 | 0.055 | 0.0312 | Small bias with slightly lower variability than the mean |

**Conclusion:**\
In clean exponential data, the **sample mean** is unbiased and has the smallest MSE, confirming it is the most efficient estimator of the true mean. Both the **median** and **trimmed mean** exhibit negative bias but are slightly more stable in variability. These results illustrate natural sampling variation and relative estimator efficiency in repeated sampling.

## Part B: Logging Error Introduces Outliers

### Scenario

To mimic recording errors, each sample contains 90% from $\mathrm{Exp}(1)$ and 10% from $\mathrm{Exp}(0.02)$ (mean $=50$), keeping $n=250$ and 1,000 replications.

### Task 1 — Generate 1,000 contaminated samples

```{r}
set.seed(6192)

rmixture <- function(n) {
  u <- runif(n)
  ifelse(u <= 0.9, rexp(n, rate = 1), rexp(n, rate = 0.02))
}

samples_B <- replicate(n_sims, rmixture(n), simplify = FALSE)
```

### Task 2

### Compute estimators for each contaminated sample

```{r}
estimates_B <- do.call(rbind, lapply(samples_B, est_stats)) |> as.data.frame()
head(estimates_B)
```

### Compute Bias, MCSD, and MSE under contamination

```{r}
mc_summary_B <- summarise_mc(estimates_B, true_mean)

kable(mc_summary_B, digits = 4,
      caption = "Part B — Bias, MCSD and MSE (90% Exp(1) + 10% Exp(0.02),
      n = 250, 1000 simulations)")

```

### Task 3 — Summary Table

| Estimator | Bias (≈) | MCSD (≈) | MSE (≈) | Interpretation Summary |
|---------------|---------------|---------------|---------------|---------------|
| **Mean** | 4.937 | 1.381 | 26.283 | Extremely inflated by outliers; very large variability and error — not robust. |
| **Median** | −0.187 | 0.082 | 0.0417 | Small negative bias; low variability — robust under contamination. |
| **Trimmed Mean (10%)** | 0.116 | 0.169 | 0.0419 | Slight bias, similar MSE to median — nearly as robust and stable. |

**Conclusion:**\

Under contaminated conditions, the **mean** becomes unreliable due to extreme bias and high MSE, while both the **median** and **10 % trimmed mean** remain accurate and consistent. The **median** performs marginally better, making it the most reliable measure of “typical” completion time when data include recording errors or outliers.

## Part C: Reflection & Questions

### 1. Estimator Performance

**Effect of outliers on average completion time**\
Outliers substantially inflate the average completion time. A few extreme values shift the mean upward, distorting the notion of a “typical” employee response.

**Performance in clean vs contaminated data**\
In clean exponential samples, the **Mean** was unbiased with the lowest MSE (≈ 0.0038).\
When 10 % extreme values were introduced, the **Mean** became unreliable (Bias ≈ +5, MSE ≈ 26), whereas the **Median** and **Trimmed Mean (10 %)** maintained small bias (≈ –0.19 and 0.12) and low MSE (≈ 0.042).\
Thus, both robust estimators continued to reflect realistic central behaviour.

**Most robust estimator**\
The **Median** proved most robust, showing negligible change in bias or MSE between clean and contaminated conditions.

**Simulation evidence supporting the conclusion**\
Across 1000 replications, the Median’s Bias and MSE remained nearly constant, while the Mean’s values increased dramatically—empirical confirmation of robustness.

### **2. Relative Efficiency**

**Most efficient estimator in clean data**\
The **Mean** was most efficient in the clean setting, achieving the smallest MSE and MCSD because it uses all sample information optimally when no outliers exist.

**Change in efficiency under contamination**\
Yes. With contamination, efficiency shifted to the **Median** and **Trimmed Mean (10%)**, whose MSEs (≈ 0.042) were far smaller than the Mean’s (≈ 26). Their robustness outweighed the Mean’s theoretical efficiency.

### **3. Real-World Implications**

**Factors for choosing between estimators**\
Analysts should evaluate data quality, distribution shape, and risk of extreme values. Right-skewed or error-prone data favour robust estimators (Median / Trimmed Mean), while validated symmetric data justify the Mean.

**Recommendation for the IT security team**\
Report the **Median completion time** as the primary KPI—robust to recording errors and extreme delays. Complement it with the **90th percentile** and **IQR** to describe variability. For validated clean datasets, include the **Mean** or **10 % Trimmed Mean** as supporting indicators.

### **Summary Reflection**

These simulations show that estimator choice must depend on data conditions.\
The **Mean** is optimal for clean, well-behaved samples, whereas the **Median** and **Trimmed Mean** safeguard reliability when contamination or long-tail behaviour is present.\
Selecting robust estimators ensures more accurate, transparent, and equitable reporting of employee-training outcomes.

# Section 2

## Part A: Building the Model with Maximum Likelihood

### **Problem Statement**

Build a logistic model that predicts whether a country’s life expectancy is above the global median using log(GDP per capita) as the predictor. Implement maximum likelihood by code (via `optim()`), verify with `glm()`, and briefly compare the estimates.

```{r}
#| label: sec2-partA
#| message: false
#| warning: false

# ---- Data and model fitting ----

library(dplyr)
library(ggplot2)
library(gapminder)

# Prepare data (2007 only)

data <- gapminder %>%
filter(year == 2007) %>%
mutate(
high_lifeExp = ifelse(lifeExp > median(lifeExp), 1, 0),
log_gdp      = log(gdpPercap)
)

y <- data$high_lifeExp
x <- data$log_gdp
n <- length(y)

# Log-likelihood for logistic regression: sum { y*eta - log(1+exp(eta)) }

loglik_logistic <- function(par, y, x) {
beta0 <- par[1]; beta1 <- par[2]
eta <- beta0 + beta1 * x
sum(y * eta - log1p(exp(eta)))
}
neg_loglik_logistic <- function(par, y, x) -loglik_logistic(par, y, x)

# MLE via optim (BFGS); request Hessian for Part B

fit_optim <- optim(
par = c(0, 0),
fn  = neg_loglik_logistic,
y = y, x = x,
method  = "BFGS",
hessian = TRUE,
control = list(reltol = 1e-10)
)
beta_hat_optim <- setNames(fit_optim$par, c("beta0","beta1"))

# Same model with glm()

fit_glm <- glm(high_lifeExp ~ log_gdp, data = data, family = binomial(link = "logit"))
beta_hat_glm <- coef(fit_glm)

# Side-by-side comparison

round(cbind(optim = beta_hat_optim, glm = beta_hat_glm), 5)

```

**Short explanation & interpretation**

`optim()` and `glm()` give virtually identical estimates: $\hat\beta_0 \approx -20.3824$ (vs $-20.3823$ from `glm`) and $\hat\beta_1 \approx 2.3489$. This agreement shows the manual MLE matches the built-in logistic regression. Because $\hat\beta_1>0$, higher $x=\log(\text{GDP per capita})$ is associated with a **higher probability** of being above the global median life expectancy.

## Part B: Estimating Uncertainty with Fisher Information

### **Problem Statement**

Quantify the precision of the fitted coefficients using the observed Fisher information from the `optim()` fit. Report standard errors and construct manual 95% Wald confidence intervals for the intercept and slope.

```{r}
#| label: sec2-partB
#| message: false
#| warning: false

# ---- Fisher information, SEs, Wald CIs ----

# Observed Fisher at MLE for logistic model equals Hessian of -logLik

Fisher_hat <- fit_optim$hessian

# Variance-covariance and SEs

vcov_hat <- solve(Fisher_hat)
se_hat   <- sqrt(diag(vcov_hat))
se_hat

# 95% Wald CIs

z <- qnorm(0.975)
beta_hat <- setNames(fit_optim$par, c("beta0","beta1"))
CI <- cbind(
lower = beta_hat - z * se_hat,
est   = beta_hat,
upper = beta_hat + z * se_hat
)
round(CI, 4)

```

**Short explanation & interpretation**\
Precision from the observed Fisher information (Hessian of the negative log-likelihood):

$$
\mathrm{SE}(\hat\beta_0)\approx 3.3848, \qquad
\mathrm{SE}(\hat\beta_1)\approx 0.3863.
$$

95% Wald confidence intervals:

$$
\beta_0 \in [-27.0166,\,-13.7483], \qquad
\beta_1 \in [1.5917,\,3.1061].
$$

**What this shows**

\- The CI for $\beta_1$ **excludes 0** with a wide margin → clear positive association between income and being above the median life expectancy.

\- Small SE for $\hat\beta_1$ indicates **good precision**.

## Part C: Visualising Parameter Stability with Profile Likelihood

### Problem Statement

Assess the stability of the slope estimate by plotting the profile log-likelihood around its MLE. Use the plot to comment on how sharply the likelihood peaks and what that implies about uncertainty.

```{r}
#| label: sec2-partC
#| message: false
#| warning: false

# ---- Profile log-likelihood for beta1 (ONE CHUNK) ----

negLL <- function(par) neg_loglik_logistic(par, y = y, x = x)

beta1_hat <- beta_hat["beta1"]
se1       <- sqrt(vcov_hat[2,2])

grid_beta1 <- seq(beta1_hat - 4*se1, beta1_hat + 4*se1, length.out = 200)

prof_vals <- numeric(length(grid_beta1))
for (k in seq_along(grid_beta1)) {
b1 <- grid_beta1[k]

# Optimize beta0 only (beta1 fixed). Wide, safe interval for logistic models:

f_b0 <- function(b0) negLL(c(b0, b1))
opt_b0 <- optimize(f_b0, c(-20, 20))
prof_vals[k] <- -opt_b0$objective
}
prof_vals <- prof_vals - max(prof_vals)  # center at 0

prof_df <- data.frame(beta1 = grid_beta1, pll = prof_vals)
ggplot(prof_df, aes(beta1, pll)) +
geom_line() +
geom_vline(xintercept = beta1_hat, linetype = 2) +
labs(
title = "Profile Log-Likelihood for β₁ (log GDP)",
x = expression(beta[1]),
y = "Profile log-likelihood (centered)"
)

```

**Short explanation & interpretation**\
The profile log-likelihood for the slope peaks near the MLE and falls off to the right:

$$
\hat\beta_1 \approx 2.3489 \quad \text{(profile peak); \; curve is sharply peaked}.
$$

**What this shows**

\- A **peaked** profile indicates $\beta_1$ is **well identified** (low uncertainty).

\- This visual check agrees with Part B’s SE ($\approx 0.386$) and narrow CI.

## Part D: Reflection & Questions

### Problem Statement

Translate the fitted model into plain language using odds ratios and practical effect sizes. Summarise the main finding for 2007 and note policy relevance and limitations (association, not causation)..

```{r}
#| label: sec2-partD
#| message: false
#| warning: false
library(tibble)

summary_tbl <- tibble(
  term     = c("Intercept (beta0)", "log(GDP) (beta1)"),
  estimate = c(as.numeric(beta_hat["beta0"]), as.numeric(beta_hat["beta1"])),
  SE       = c(as.numeric(se_hat["beta0"]),   as.numeric(se_hat["beta1"])),
  OR       = exp(c(as.numeric(beta_hat["beta0"]), as.numeric(beta_hat["beta1"])))
)

summary_tbl %>% mutate(across(where(is.numeric), ~round(.x, 4)))


```

**Short explanation & interpretation**\
Effect size on the odds scale:

$$
\text{OR}=\exp(\hat\beta_1)=\exp(2.3489)\approx 10.5.
$$

Useful translations:

$$
\text{10\% GDP increase: } \exp(0.1\times 2.3489)\approx 1.26 \ (\sim 26\%\ \text{higher odds}).
$$

$$
\text{Doubling GDP: } \exp((\ln 2)\times 2.3489)\approx 5.1 \ (\sim 5\times\ \text{higher odds}).
$$

**What this shows** - The effect is **large and practically meaningful**. - Combined with small SEs and a peaked profile, the estimate is **precise** for 2007.

### Section 2: Overall Conclusion

We model the probability of being above the global median life expectancy with a logistic link:

$$
\operatorname{logit}\!\big(\Pr(Y=1\mid x)\big)=\beta_0+\beta_1 x, 
\qquad x=\log(\mathrm{GDP\ per\ capita}).
$$

**Findings**

\- **Direction & size:** $\hat\beta_1\approx 2.3489>0$; on odds, $\exp(\hat\beta_1)\approx 10.5$\
→ 10% GDP rise $\approx 26\%$ higher odds; doubling GDP $\approx 5\times$ higher odds.

\- **Uncertainty:** $\mathrm{SE}(\hat\beta_1)\approx 0.386$; 95% CI $[1.5917,\,3.1061]$ **excludes 0**.

\- **Stability:** Profile likelihood is **sharply peaked** at $\hat\beta_1$ → **precise** slope estimate.

**Implication**\
For 2007, richer countries are **substantially more likely** to exceed the global median life expectancy—consistent with UNSDG Goal 3. *Caveat:* the analysis is cross-sectional and bivariate (associational, not causal).
